### 2021-04-19 paper/appendix_STA561_project.Rmd
GetPossiblePatterns function

---
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
header-includes: \usepackage{hyperref}
linestretch: 1
link-citations: yes
linkcolor: blue
fontfamily: mathpazo
fontsize: 12pt
---
\begin{flushright} 
	\end{flushright}
	\begin{center} \textbf{Parlr: Parallelized Record Linkage in R}

@@

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Most methods for conducting record linkage, the task of identifying duplicate records across datasets, are derived from the foundational article by Fellegi and Sunter in 1969. Within this family of methods, two methods have notable strengths: `fastlink`, proposed by Enamorado et al (2019), continues a is able to conduct linkage on incredibly large datasets

@@

Our notation and assumptions closely follow that of Sadinle (2017). Denote two files as $A$ and $B$, with $n_A$ and $n_B$ records respectively, and with records indexed as $i \in \{1, \ldots, n_A\}$ in $A$ and $j \in \{1, \ldots, n_B\}$ in $B$. Without loss of generality, label the files such that $n_A \geq n_B$. We also assume there are no duplicates within files, only across. For each record pair under consideration, we generate a comparison vector $\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}$, where  $F$ is the number of fields used in the linkage and each $\gamma_{ij}^f$ takes on a value $l \in \{1, \ldots, L_f\}$ indicating level agreement between the two records on a specified field.

@@

Wherever possible, we reserve superscripts for denoting field and level, while reserving subscripts for record indices. For example, $\mathbf{m}^f = (m^{f1}, \ldots, m^{fL_f})$ is the probability distribution governing field $f$ for matching records, and $\mathbf{m}_{ij}= \prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(m^{fl}\right)^{\mathbf{1}_{\gamma_{ij}^f = l}} = P(\boldsymbol{\gamma}_{ij}|Z_j = i)$ is product of the relevant of the appropriate $\mathbf{m}$ parameters for record pair $(i,j)$. We hope that these conventions avoid overloaded notation in the likelihood and subsequent derivations.

@@

(This will be revised to talk more explicitly about sampling $Z_j$ independently)

Our model differs from that of Sadinle 2017 through its explicit dependence on a beta random variable $\lambda$ that models the rate of matching across records. Sadinle marginalizes over such a random variable in his derivations of  the "beta prior for bipartite matching," but here we provide derivations without marginalizing in order to specify differing rates of matching for different linkage clusters.

## Prior Distributions and Likelihood

@@

@@

Broadly speaking, we increase our computational efficiency by recognizing that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. Let $H$ be the set of unique agreement patterns in the data, let $P$ denote the total number of unique agreement patterns. Note that $P$ is bounded above by $\prod_{f=1}^F L_f$, and that this bound does not scale with $n_A$ or $n_B$. We index these agreement patterns by $p \in \{1, \ldots, P\}$, and say $(i,j) \in h_p$ when the $(i,j)$ pair exhibits the $p^{th}$ agreement pattern. Wherever possible, we conduct calculations over these $P$ agreement patterns rather than the $n_A \times n_B$ record pairs.

@@

We store this information in a nested list $\tilde{\Gamma}$ where the $p^{th}$ component of the $j^{th}$ list contains a vector of records in $A$ that share agreement pattern $h_p$ with record $j \in B$. For each $p$, we also calculate $|h_p| = \sum_{i=1}^{n_A}\sum_{j=1}^{n_B} \mathbf{1}_{(i,j) \in h_p}$ total instances of agreement pattern $h_p$ throughout the data, and also for each $j$, we  calculate $|h_p|_j = \sum_{i=1}^{n_A} \mathbf{1}_{(i,j) \in h_p}$ the instances of agreement pattern $p$ among the comparison vectors between record $j \in B$ and each of the $n_A$ records in $A$.

For large data, we can partition the two datasets $A$ and $B$ into smaller blocks $\{A_m\}$ and $\{B_m\} for more manageable computations. On a single machine, we can read-in data sequentially, conduct hashing, collect results, and delete the original data from memory before continuing with the next chunk of data. With multiple cores or multiple machines, this can be done in parallel. Storing this hashed information still becomes burdensome for large data, but this hashing method greatly expands the capabilities of the Fellegi-Sunter framework.

@@

After receiving matching statuses from $\mathbf{Z}$, the Sadinle method calculates $\alpha_{fl}(\mathbf{Z})$ and $\beta_{fl}(\mathbf{Z})$ for each field and level. This constitutes $2 \times \sum L_f$ many summations over $n_A \times n_B$ quantities, and becomes computationally burdensome with large data. In contrast, we recognize that each unique agreement pattern contributes to the posterior $\alpha(\mathbf{Z})$ and $\beta(\mathbf{Z})$ vectors in the same way. In fact, if we denote $|h_p^m| = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}$ to be the number of matching record pairs with agreement pattern $h_p$, then the contribution of pairs of pattern $h_p$ to the $\alpha(\mathbf{Z})$ vector is simply $\mathbf{Z}(h_p) \times h_p$. Thus our posterior update for $\alpha$ is simply $\alpha(\mathbf{Z}) = \alpha_0 + \sum_{p=1}^P |h_p^m| \times h_p$. Then, we can easily calculate $|h_p^u|$, the number of nonmatching record pairs of agreement pattern $p$, by subtracting the number of matching pairs from the total present in the data; that is $|h_p^u| = |h_p| - |h_p^m|$. From this, we can update our $\beta$ parameter through  $\beta(\mathbf{Z}) = \beta_0 + \sum_{p=1}^P |h_p^u| \times h_p$. Note that these constitute $P$ many summations over $n_B$ quantities, and thus avoid the $n_A \times n_B$ summation from the original method.

@@

Sadinle uses a prior for $\mathbf{Z}$ that induces the a full conditional for $Z_j$ that strictly enforces one-to-one matching. Particular, this sampler removed previously matches records from the set of candidate records when sampling $Z_j$, creating a dependency that makes the sampler *inherently serial*.By weakening the one-to-one requirement, our full conditional for $Z$ does not depend on the rest of the $\mathbf{Z}_{-j}$ vector, and thus can be computed in parallel. More importantly, since only the agreement pattern of $Z_j$ is used for calculations within the Gibbs sampler, and not the particular record label, we can conduct this sampling only at the level of the unique agreement patterns. This boosts computation time far greater than parallelization.

To do this we calculate the Fellegi Sunter weight $w_{h_p}$ for each unique pattern,  sample the agreement pattern between $j$ and its potential match, and then sample the record label uniformly among viable records. More concretely, define $H(Z_j)$ to be the agreement pattern between $j$ and its potential match, and say $H(Z_j) = h_{P+1}$ when $Z_j = n_A + 1$. Then,

$$H\left(Z_j^{(s+1)}\right) | \mathbf{m}, \mathbf{u}, \mathbf{Z^{(s)}} \propto
\begin{cases} 
    w_{h_p}\times |h_p|_j  & h(z_j) \in H; \\
     n_A \frac{n_B - D + \beta_{\lambda}}{D + \alpha_{\lambda}} &  z_j  = n_A + 1 \\
\end{cases}$$
<!-- $$p\left(Z_j^{(s+1)} = i\right) = \frac{1}{\left|H\left(Z_j^{(s+1)}\right)\right|}; \;\;i \text{ such that } (i, j) \in H\left(Z_j^{(s+1)}\right)$$ -->

@@

## Enforcing Bipartite Matching

Add this

@@

- Replicating the Sadinle Study

- `parlr` and `BRL` speed comparison

# Data Analysis

- Necessary? El Salvador data? Or perhaps needs to be a bigger dataset?

# Discussion

We note that this often not a reasonable assumption; for example if we know that two records differ on gender, this does not give any information on whether their year of births are the same, but it does suggest that its more likely that their first names are different.

>>>>

---
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
header-includes: \usepackage{hyperref}
linestretch: 1
link-citations: yes
linkcolor: blue
fontfamily: mathpazo
fontsize: 12pt
---
\begin{flushright} 
	\end{flushright}
	\begin{center} \textbf{Parlr: Parallelized Record Linkage in R} \linebreak
	\textbf{Technical Appendix}

@@

	April 23, 2021

	STA 561

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<!-- ## Introduction -->

<!-- Most methods for conducting record linkage, the task of identifying duplicate records across datasets, are derived from the foundational article by Fellegi and Sunter in 1969. Within this family of methods, two methods have notable strengths: `fastlink`, proposed by Enamorado et al (2019), continues a is able to conduct linkage on incredibly large datasets -->

@@

Denote two files as $A$ and $B$, with $n_A$ and $n_B$ records respectively, with records indexed as $i \in \{1, \ldots, n_A\}$ in $A$ and $j \in \{1, \ldots, n_B\}$ in $B$. Without loss of generality, label the files such that $n_A \geq n_B$. We also assume there are no duplicates within files, only across.

For each record pair under consideration, we generate a comparison vector $\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}$, where  $F$ is the number of fields used in the linkage and each $\gamma_{ij}^f$ takes on a value $l \in \{1, \ldots, L_f\}$ indicating the level agreement between the two records on a specified field. For ease of modeling and computation, we restrict these similarity scores to be discrete, ordinal variables, and the construction of these is left to the modeler. With notation $\gamma_{ij}^f \in \{0, \ldots, L_f - 1\}$ for the number of possible agreement levels for field $f$, we use binary 0-1 variables to indicate exact matching, and 0-1-2 variables to provide an option for partial matching. For text data, we calculate similarity based on Levenstein distance or some other text similarity score, and bin these scores to integers for use in the model.

@@

Wherever possible, we reserve superscripts for denoting field and level, while reserving subscripts for record indices. For example, $\mathbf{m}^f = (m^{f1}, \ldots, m^{fL_f})$ is the probability distribution governing field $f$ for matching records, and $\mathbf{m}_{ij}= \prod_{f=1}^{F}\prod_{l=1}^{L_f} \left(m^{fl}\right)^{\mathbf{1}_{\gamma_{ij}^f = l}} = P(\boldsymbol{\gamma}_{ij}|Z_j = i)$ is the product of the appropriate $\mathbf{m}$ parameters for record pair $(i,j)$. We hope that these conventions avoid overloaded notation in the likelihood and subsequent derivations.

@@

<!-- Our model differs from that of Sadinle 2017 through its explicit dependence on a beta random variable $\lambda$ that models the rate of matching across records. Sadinle marginalizes over such a random variable in his derivations of  the "beta prior for bipartite matching," but here we provide derivations without marginalizing in order to specify differing rates of matching for different linkage clusters.  -->

<!-- ## Prior Distributions and Likelihood -->

@@

Since our Gibbs procedure does not strictly enforce one-to-one matching, it is possible for the final Bayes estimate to link multiple records in $B$ to one record in $A$. The modeler can either report both such matches (with their respective posterior match probabilities), or resolve these conflicts by accepting only the match with highest posterior probability. Rigorous proof of this resolution procedure is deferred to future work.

@@

Broadly speaking, we increase our computational efficiency by recognizing that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. Let $\mathcal{H}$ be the set of unique agreement patterns in the data, let $P$ denote the total number of unique agreement patterns. Note that $P$ is bounded above by $\prod_{f=1}^F L_f$, and that this bound does not scale with $n_A$ or $n_B$. We index these agreement patterns by $p \in \{1, \ldots, P\}$, and say $(i,j) \in h_p$ when the $(i,j)$ pair exhibits the $p^{th}$ agreement pattern. Wherever possible, we conduct calculations over these $P$ agreement patterns rather than the $n_A \times n_B$ record pairs.

@@

We store this information in a nested list $\tilde{\Gamma}$ where the $p^{th}$ component of the $j^{th}$ list contains a vector of records in $A$ that share agreement pattern $p$ with record $j \in B$. For each $p$, we also calculate $H_p = \sum_{i=1}^{n_A}\sum_{j=1}^{n_B} \mathbf{1}_{(i,j) \in h_p}$, the total instances of agreement pattern $p$ throughout the data, and also for each $j$, we  calculate $H_{p_j} = \sum_{i=1}^{n_A} \mathbf{1}_{{(i,j) \in h_p}}$ the instances of agreement pattern $p$ among the comparison vectors between record $j \in B$ and each of the $n_A$ records in $A$.

For large data, we can partition the two datasets $A$ and $B$ into smaller blocks $\{A_m\}$ and $\{B_m\}$ for more manageable computations. On a single machine, we can read-in data sequentially, conduct hashing, collect results, and delete the original data from memory before continuing with the next chunk of data. With multiple cores or multiple machines, this can be done in parallel. Storing this hashed information still becomes burdensome for large data, but this hashing method greatly expands the capabilities of the Fellegi-Sunter framework.

@@

After receiving matching statuses from $\mathbf{Z}$, the Sadinle method calculates $\alpha_{fl}(\mathbf{Z})$ and $\beta_{fl}(\mathbf{Z})$ for each field and level. This constitutes $2 \times \sum L_f$ many summations over $n_A \times n_B$ quantities, and becomes computationally burdensome with large data. In contrast, we recognize that each unique agreement pattern contributes to the posterior $\alpha(\mathbf{Z})$ and $\beta(\mathbf{Z})$ vectors in the same way. In fact, if we denote $H_p^m = \sum_{j=1}^{n_B} \mathbf{1}_{(Z_j, j) \in h_p}$ to be the number of matching record pairs with agreement pattern $p$, then the contribution of pairs of pattern $p$ to the $\alpha(\mathbf{Z})$ vector is simply $H_p^m \times h_p$. Thus our posterior update for $\alpha$ is simply $\alpha(\mathbf{Z}) = \alpha_0 + \sum_{p=1}^P H_p^m \times h_p$. Then, we can easily calculate $H_p^u$, the number of nonmatching record pairs of agreement pattern $p$, by subtracting the number of matching pairs from the total present in the data; that is $H_p^u = H_p - H_p^m$. From this, we can update our $\beta$ parameter through  $\beta(\mathbf{Z}) = \beta_0 + \sum_{p=1}^P H_p^u \times h_p$. Note that these constitute $P$ many summations over $n_B$ quantities, and thus avoid the $n_A \times n_B$ summation from the original method.

@@

Sadinle uses a prior for $\mathbf{Z}$ that induces the a full conditional for $Z_j$ that strictly enforces one-to-one matching. In particular, this sampler removes previously matches records from the set of candidate records when sampling $Z_j$, creating a dependency that makes the sampler *inherently serial*. By weakening the one-to-one requirement, our full conditional for $Z$ does not depend on the rest of the $\mathbf{Z}_{-j}$ vector, and thus can be computed in parallel. More importantly, since only the agreement pattern of $Z_j$ is used for calculations within the Gibbs sampler, and not the particular record label, we can conduct this sampling only at the level of the unique agreement patterns. This boosts computation time far greater than parallelization.

To do this, we calculate the Fellegi Sunter weight $w_{h_p}$ for each unique pattern,  sample the agreement pattern between $j$ and its potential match, and then sample the record label uniformly among viable records. More concretely, define $h(Z_j)$ to be the agreement pattern between $j$ and its potential match, and say $h(Z_j) = h_{P+1}$ when $Z_j = n_A + 1$. Then,

$$h\left(Z_j^{(s+1)}\right) | \mathbf{m}, \mathbf{u}, \mathbf{Z^{(s)}} \propto
\begin{cases} 
    w_{h_p}\times H_{p_j}  & p \leq P; \\
     n_A \frac{n_B - D + \beta_{\lambda}}{D + \alpha_{\lambda}} &   p = P + 1 \\
\end{cases}$$
<!-- $$p\left(Z_j^{(s+1)} = i\right) = \frac{1}{\left|H\left(Z_j^{(s+1)}\right)\right|}; \;\;i \text{ such that } (i, j) \in H\left(Z_j^{(s+1)}\right)$$ -->

@@

@@

In his 2017 paper, Sadinle demonstrated the strength of his method by running `BRL` on simulated pairs of files with differing levels of errors and overlap. Comparing our method against `BRL` on these same simulated datasets, we find that our method only has weakened performance in the extreme scenario of very high errors and very high overlap across files. We use the following measures of accuracy in our analysis: although using only recall and precision is common in the literature, we find it useful to have a single metric the gauge overall performance, and thus also include F measure.

$$\text{Recall} = \frac{\text{Matches Correctly Identified}}{\text{True Matches in Data}}$$
$$\text{Precision} = \frac{\text{Declared Matches}}{\text{True Matches in Data}}$$
$$\text{F-Measure} = 2\left(\frac{\text{Recall} \times \text{Precision}}{\text{Recall} + \text{Precision}}\right)$$

```{r fig.align='center'}
knitr::include_graphics("../notes/figures/sadinle_sim_plot.png")
```

To demonstrate speed, we generate comparison vectors from pre-specified distributions so that we can easily increase the size of the linkage problem. Distributions are meant to emulate the behavior of similarity scores across first name, last name, and day, month, and and year of birth, with 1 indicating an exact match on a particular field and 0 indicating a nonmatch. We simulate these data for different values of $n_A$ and $n_B$, and compare the runtime of `parlr` against `BRL`. We see that at low data size, `BRL` outperforms, but that `parlr` is significantly faster at handling larger data.

In particular, runtime for `BRL` seems to grow quadratically while runtime for `parlr` seems to grow linearly.

\begin{table}[ht]
\centering
\begin{tabular}{rrr}
  \hline
 & m & u \\ 
  \hline
fname\_0 & 0.05 & 0.99 \\ 
  fname\_1 & 0.95 & 0.01 \\ 
  lname\_0 & 0.05 & 0.99 \\ 
  lname\_1 & 0.95 & 0.01 \\ 
  day\_0 & 0.05 & 0.97 \\ 
  day\_1 & 0.95 & 0.03 \\ 
  month\_0 & 0.05 & 0.92 \\ 
  month\_1 & 0.95 & 0.08 \\ 
  year\_0 & 0.05 & 0.93 \\ 
  year\_1 & 0.95 & 0.07 \\ 
   \hline
\end{tabular}
\end{table}

<!-- ```{r fig.align='center'} -->
<!-- knitr::include_graphics("../notes/figures/speed_distributions.png") -->
<!-- ``` -->

```{r fig.align='center'}
knitr::include_graphics("../notes/figures/sadinle_speed_plot.png")
```

The above discussion suggests that for fixed $n_B$, computation time should remain mostly constant with growing $n_A$. Shockingly, this seems to be true. In the plot below, fixing $n_B = 500$, we see linear growth for the runtime under `BRL` as $n_A$ increases, with much more static runtime under `parlr`. The slight increases in runtime that we do see are due primarily to the hashing step, which again can be run in parallel for large data.

```{r fig.align='center'}
knitr::include_graphics("../notes/figures/speed_plot_fixed_nB.png")
```

We note here that `BRL` is coded in C, which makes for unfair comparison against `parlr`, currently only built in R. Additionally, although `parlr` is amenable to parallezation, this simulation was run on a single core. Running `parlr` in C++ with paralellization for the hashing step and sampling the matching status of the record pairs should lead to even more drastic results.
